A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-15 paper/fascd.tex
work on NCP residual text; we need a both-bounds test case

@@

In this section we consider several practical concerns regarding the implementation and application of Algorithm \ref{alg:fascd}.  We address convergence criterion, smoothers and coarse solvers, and finally the F-cycle (sometimes called \emph{full multigrid cycle}) extension.

Finite-dimensional, box-constrained FE VI problem \eqref{eq:fe:vi} is equivalent to a \emph{mixed complementarity problem} (MCP) \cite{FacchineiPang2003}, a strong form whose statement may be clarifying.  Denote $x_p$ for a node of the finest-level mesh $\mathcal{T}^J$ and $\psi_p(x)$ for the corresponding hat function.  Let $\mathcal{X}^J=\big\{w\in\cV^J\,:\,w(x_p)=g_D^J(x_p) \text{ if } x_p\in \partial_D\Omega\big\}$, thus $\cK^J = \left\{\underline{\gamma}^J \le w \le \overline{\gamma}^J\right\} \cap \mathcal{X}^J$.  If $m_D$ is the number of nodes $x_p$ in the Dirichlet boundary, let us define $I_D$ as the $m_D\times m_J$ matrix which extracts Dirichlet nodal values; its transpose $I_D^\top$ injects Dirichlet boundary degrees of freedom into $(\mathcal{V}^J)'$.  Let us use the symbols $a\perp b$ to indicate that $a,b \in \RR^n$ are \emph{complementary}, that is, $a\ge 0$ and $b\ge 0$ and $ab = 0$ (entrywise).  Problem \eqref{eq:fe:vi}, expressed as VI$\left(f^J(\cdot)-\ell^J,\left[\,\underline{\gamma}^J,\overline{\gamma}^J\right] \cap \mathcal{X}^J\right)$ in the operations research literature \cite{FerrisPang1997}, is equivalent to the following MCP with additional equality constraints:
\begin{align}
\underline{\gamma}^J \le u^J \le \overline{\gamma}^J \quad &\perp \quad f^J(u^J) - \ell^J - I_D^\top \lambda \label{eq:fe:mcp} \\
I_D u^J &= g_D^J\big|_{\partial_D\Omega} \notag
\end{align}
for some Lagrange multipliers $\lambda \in \RR^{m_D}$.

MCP \eqref{eq:fe:mcp} says that exactly one of five conditions holds for the solution $u^J$ at $x_p \in \mathcal{T}^J$.  If $x_p \notin \partial_D\Omega$ then one of these four conditions hold:
\begin{itemize}
\item $\underline{\gamma}^J(x_p)<u^J(x_p)<\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} = 0$,
\item $\underline{\gamma}^J(x_p)=u^J(x_p)<\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} \ge 0$,
\item $\underline{\gamma}^J(x_p)<u^J(x_p)=\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} \le 0$,
\item $\underline{\gamma}^J(x_p)=u^J(x_p)=\overline{\gamma}^J(x_p)$.
\end{itemize}
Otherwise, if $x_p \in \partial_D\Omega$ then
\begin{itemize}
\item $u^J(x_p)=g_D^J(x_p)$ and $\lambda_p=\ip{f^J(u^J)-\ell^J}{\psi_p}$.
\end{itemize}
The cases might be labeled \emph{inactive}, \emph{lower active}, \emph{upper active}, \emph{pinched}, and \emph{boundary}, respectively.  Where case \emph{pinched} applies, $\ip{f^J(u^J)-\ell^J}{\psi_p}$ can have any real value.

Because the problem is inequality-constrained, repeated application of \pr{fascd-vcycle} to solve VI \eqref{eq:fe:vi} will not generally reduce the residual $f^J(w^J) - \ell^J$ to zero everywhere.  However, MCP \eqref{eq:fe:mcp} identifies candidate quantities which should go to zero, and which we can compare to a tolerance.  FIXME first, for $w^J \in \mathcal{K}^J$ we define the finest-level \emph{nodal complementarity residual} in $\RR^{m_J}$.  It is defined in the four cases listed after \ref{eq:fe:mcp}, namely inactive, lower active, upper active, and boundary:
\begin{equation}
\hat r(w^J)_p = \begin{cases}
    \ip{f^J(w^J)-\ell^J}{\psi_p}, & \underline{\gamma}^J(x_p) < w^J(x_p) < \overline{\gamma}^J(x_p), \\
    \min\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \underline{\gamma}^J(x_p), \\
    \max\{\ip{f^J(w^J)-\ell^J}{\psi_p},0\}, & w^J(x_p) = \overline{\gamma}^J(x_p), \\
    w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega. \end{cases} \label{eq:cpresidual}
\end{equation}
Where the box constraints pinch at $x_p \notin \partial_D\Omega$, i.e.~where $\underline{\gamma}^J(x_p)=u^J(x_p)=\overline{\gamma}^J(x_p)$, we define $\hat r(w^J)_p=0$.  Note that $\hat r:\mathcal{V}^J \to \RR^{m_J}$ is a computable nonlinear map derived from $f^J$ and the vectors $\ell^J$, $\underline{\gamma}^J$, and $\overline{\gamma}^J$.  For unconstrained PDEs one would use only the inactive and boundary cases.

The idea here is that $\hat r(w^J)$ is nonzero only if there are nodes where MCP \eqref{eq:fe:mcp} is violated, either because the pointwise residual $\ip{f^J(w^J)-\ell^J}{\psi_p}$ remains nonzero where the constraints are inactive, or because it has the wrong sign on an actively-constrained degree of freedom, or because the Dirichlet boundary condition is not satisfied.

FIXME Fischer-Burmeister semismooth alternative for convergence

@@

FIXME  Recall that a $j$th-level hat function based at node $x_p^j$ is denoted $\psi_p^j$, for an iterate $g^j$ and an admissible correction $y^j \in \mathcal{D}^j$ we define the \emph{downward nodal residual} as
\begin{equation}
(\hat r_{\text{down}}(y^j))_p = \begin{cases} \ip{f^j(g^j+y^j)-\ell^j}{\psi_p^j}, & y^j(x_p^j) > \phi^j(x_p^j), \\
                                  \min\{\ip{f^j(g^j+y^j)-\ell^j}{\psi_p^j},0\}, & y^j(x_p^j) = \phi^j(x_p^j). \end{cases} \label{eq:dncpresidual}
\end{equation}
The \emph{upward nodal residual} $\hat r_{\text{up}}(z^j)$ for $z^j \in \mathcal{U}^j$ is similarly defined using $\chi^j$.

FIXME smoothers including semi-smooth and active set \cite{BensonMunson2006}

>>>>

\newcommand{\rNC}{r_{\text{NC}}}
\newcommand{\rSS}{r_{\text{SS}}}

@@

In this section we consider several practical concerns when implementing and applying \pr{fascd-vcycle} Algorithm \ref{alg:fascd}.  We address convergence criteria, smoothers, and coarse solvers, plus the F-cycle extension of the Algorithm, sometimes called a \emph{full multigrid cycle}.

Because the problem is inequality-constrained, repeated application of \pr{fascd-vcycle} to solve the finite-dimensional, box-constrained VI \eqref{eq:fe:vi} will not generally reduce the residual $f^J(w^J) - \ell^J$ to zero everywhere.  However, \eqref{eq:fe:vi} is equivalent to a \emph{mixed complementarity problem} (MCP) \cite{FacchineiPang2003}, a strong form amenable to convergence criteria.  Let $x_p \in \mathcal{T}^J$ be a finest-level node, and $\psi_p(x)$ the corresponding hat function.  Let $\mathcal{X}^J=\big\{w\in\cV^J\,:\,w(x_p)=g_D^J(x_p) \text{ if } x_p\in \partial_D\Omega\big\}$, thus $\cK^J = \left\{\underline{\gamma}^J \le w \le \overline{\gamma}^J\right\} \cap \mathcal{X}^J$.  If $m_D$ is the number of nodes $x_p$ in the Dirichlet boundary, let $I_D$ be the $m_D\times m_J$ matrix which extracts Dirichlet nodal values; its transpose $I_D^\top$ injects Dirichlet boundary degrees of freedom into $(\mathcal{V}^J)'$.  Finally, we will write $a\perp b$ when $a,b \in \RR^n$ are \emph{complementary}, that is, when $a_i \ge 0$ and $b_i \ge 0$ and $a_i b_i = 0$ for $i=0,\dots,n-1$.  Then problem \eqref{eq:fe:vi}, which would be expressed VI$\left(f^J(\cdot)-\ell^J,\left[\,\underline{\gamma}^J,\overline{\gamma}^J\right] \cap \mathcal{X}^J\right)$ in the operations research literature \cite{FerrisPang1997}, is equivalent to the following MCP for the solution $u^J \in \mathcal{K}^J$, but with additional equality constraints:
\begin{align}
\underline{\gamma}^J \le u^J \le \overline{\gamma}^J \quad &\perp \quad f^J(u^J) - \ell^J - I_D^\top \lambda \label{eq:fe:mcp} \\
I_D u^J &= g_D^J\big|_{\partial_D\Omega} \notag
\end{align}
for some Lagrange multipliers $\lambda \in \RR^{m_D}$.  By definition, MCP \eqref{eq:fe:mcp} says that certain node-wise conditions hold for $u^J$ at $x_p \in \mathcal{T}^J$.  If $x_p \notin \partial_D\Omega$ then one of these four conditions hold:
\begin{itemize}
\item $\underline{\gamma}^J(x_p)<u^J(x_p)<\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} = 0$,
\item $\underline{\gamma}^J(x_p)=u^J(x_p)<\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} \ge 0$,
\item $\underline{\gamma}^J(x_p)<u^J(x_p)=\overline{\gamma}^J(x_p)$ and $\ip{f^J(u^J)-\ell^J}{\psi_p} \le 0$,
\item $\underline{\gamma}^J(x_p)=u^J(x_p)=\overline{\gamma}^J(x_p)$.
\end{itemize}
Otherwise, if $x_p \in \partial_D\Omega$ then
\begin{itemize}
\item $u^J(x_p)=g_D^J(x_p)$ and $\lambda_p=\ip{f^J(u^J)-\ell^J}{\psi_p}$.
\end{itemize}
The cases might be labeled \emph{inactive}, \emph{lower active}, \emph{upper active}, \emph{pinched} (\emph{both active}), and \emph{boundary}, respectively.  Where \emph{pinched} applies, $\ip{f^J(u^J)-\ell^J}{\psi_p}$ can have any real value.

Thus the MCP \eqref{eq:fe:mcp} identifies expressions whose norms should go below tolerance.  For for $w^J \in \mathcal{K}^J$, let $r_p = \ip{f^J(w^J)-\ell^J}{\psi_p}$, the pointwise (ordinary) residual, as a notational convenience.  First, interpreting the MCP directly, we define the finest-level \emph{nodal complementarity residual} $\rNC$ in $\RR^{m_J}$ from the five cases:
\begin{equation}
\rNC(w^J)_p = \begin{cases}
    r_p, & \underline{\gamma}^J(x_p) < w^J(x_p) < \overline{\gamma}^J(x_p), \\
    \min\{r_p,0\}, & w^J(x_p) = \underline{\gamma}^J(x_p), \\
    \max\{r_p,0\}, & w^J(x_p) = \overline{\gamma}^J(x_p), \\
    0, & \underline{\gamma}^J(x_p)=w^J(x_p)=\overline{\gamma}^J(x_p), \\
    w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega. \end{cases} \label{eq:rNC}
\end{equation}
(The first four cases assume $x_p \notin \partial_D\Omega$.)  Note that $\rNC : \mathcal{K}^J \to \RR^{m_J}$ is a computable nonlinear map defined from $f^J$ and the vectors $\ell^J$, $\underline{\gamma}^J$, and $\overline{\gamma}^J$, and that for PDEs without constraints one uses only the first and last cases.  In summary, $\rNC(w^J)_p$ is nonzero at nodes where MCP \eqref{eq:fe:mcp} is violated, either because the pointwise residual $\ip{f^J(w^J)-\ell^J}{\psi_p}$ remains nonzero where the constraints are inactive, or because it has the wrong sign at an actively-constrained degree of freedom, or because the boundary condition is not satisfied.

As a practical matter, however, direct implementation of \eqref{eq:rNC} is sensitive to rounding error because $\rNC$ is a discontinuous function of $w^J(x_p)$.  For example, if $w^J(x_p)$ exceeds $\underline{\gamma}^J(x_p)$ by $10^{-16}$, then, instead of returning the minimum of $z_p = \left<f^J(w^J)-\ell^J,\psi_p\right>$ and zero (the second case), the first case applies and $\rNC(w^J)_p = z_p$, which might be large (positive).  This would suggest that the iteration is far from converging.  On the other hand, an advantage of formula \eqref{eq:rNC} is that it applies as stated if $\underline{\gamma}^J(x_p)=-\infty$ or $\overline{\gamma}^J(x_p)=+\infty$; compared to \eqref{eq:rSS} below, the difference $w^J(x_p) - \underline{\gamma}^J(x_p)$, for example, is never computed.

%\newcommand{\phiFB}{\phi_{\text{FB}}}
We will instead measure convergence using a \emph{semi-smooth residual} based on the Fischer-Burmeister function \cite{BensonMunson2006,Ulbrich2011} which is defined as follows for $a,b\in\RR$:
\begin{equation}
\phi(a,b) = a + b - \sqrt{a^2 + b^2}. \label{eq:phiFB}
\end{equation}
Observe that $\phi(a,b)=0$ if and only if $a,b$ are complementary, and also that $\phi$ semi-smooth (and continuous).  Now let
\begin{equation}
\rSS(w^J)_p = \begin{cases}
r_p, & \underline{\gamma}^J(x_p) = -\infty \text{ and }\overline{\gamma}^J(x_p) = +\infty, \\
\phi\left(w^J(x_p) - \underline{\gamma}^J(x_p), r_p\right), & \overline{\gamma}^J(x_p) = +\infty, \\
\phi\left(\overline{\gamma}^J(x_p) - w^J(x_p), -r_p\right), & \underline{\gamma}^J(x_p) = -\infty, \\
\text{FIXME: this case needs testing} \\
w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega.
\end{cases} \label{eq:rSS}
\end{equation}

@@

FIXME smoother semantics; shift ``$w^j+y^j$'' etc.

FIXME smoothers including Newton active set \cite{BensonMunson2006} or Newton semi-smooth \cite{BensonMunson2006}, with either direct linear solve or a fixed number of Krylov steps
